Universal relations after GW170817

### 2018-11-28 paper/bin-love-after-gw170817.tex
Comments by Katerina

\definecolor{red(ncs)}{rgb}{0.77, 0.01, 0.2}
\newcommand{\ny}[1]{\textcolor{blue}{NY: #1} }

@@

In the current analysis, we consider only EoSs which have been constrained by the recent binary neutron star merger event, GW170817.
Important recent work~\cite{LIGO:posterior} uniformly sampled EoS parameters in order to derive a marginalized posterior on the pressure as a function of mass density, as seen in Fig. 2 of~\cite{LIGO:posterior}.
In our work, we similarly sample spectral coefficients within the ranges: $\gamma_0 \in \lbrack 0.2,2 \rbrack$, $\gamma_1 \in \lbrack -1.6,1.7 \rbrack$, $\gamma_2 \in \lbrack -0.6,0.6 \rbrack$, $\gamma_0 \in \lbrack -0.02,0.02 \rbrack$, and further restricting the adiabatic index to be $\Gamma \in \lbrack 0.6,4.5 \rbrack$, ensuring our parameterization exposes a wide range of viable EoSs~\cite{Lindblom:parameters}.
In addition, we impose similar constraints as was done in~\cite{LIGO:posterior}: (i) causality within 10\%, and (ii) EoS priors must support NS masses up to $1.97 \text{ M}_{\odot}$, consistent with astrophysical observations.
We then utilize the posterior defined by GW170817 by rejecting any EoS sample which does not lie within the 90\% credible posterior region, resulting in a sample of 35 EoSs shown in Fig.~\ref{fig:eos}.
Following~\cite{Read2009}, the parameterized high-density core EoSs generated above are matched to the low-density crust EoS Sly~\cite{Douchin:2001sv} at about half of the nuclear saturation density, $\rho_{\text{stitch}}=1.3 \times 10^{14} \text{ g/cm}^3$.
For comparison, we also include a second sample of randomly generated EoSs, without restricting to GW170817s posterior. {\ny{It would be good to make this figure a double pane one, so that you can show the constrained set and the unconstrained set side-by-side.}}
\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{EoSs.eps}
\end{center}
\caption{
Representative set of EoSs used in our analysis. 
These were populated by uniformly sampling the spectral coefficient parameter space, then restricting to the 90\% credible level shown in cyan~\cite{LIGO:posterior}.
In addition, the following constraints were also applied: (i) the EoS must have a causal structure within 10\%, and (ii) the EoS must support maximum NS mass of at least $1.97 \text{ M}_{\odot}$, consistent with astrophysical observations.
Finally, the high density core EoSs generated above were stitched to the low-density crust EoS of SLy~\cite{Douchin:2001sv} at $\rho_{\text{stitch}}=1.3 \times 10^{14} \text{ g/cm}^3$.
}
\label{fig:eos}
\end{figure}

@@

\subsection{Binary love relations}\label{sec:binary}
Next we consider improvements to the binary Love universal relations.
In particular, we consider two distinct classes of NSs: nuclear matter EoSs and hybrid quark-hadron star EoSs~\cite{Paschalidis2018,Alford:2017qgh,1971SvA....15..347S,Zdunik:2012dj,Alford:2013aca}.
In the latter, the low-density nucleonic crust region of the EoS experiences a first-order phase transition into a high density quark matter core at a given transition pressure $P_{\text{tr}}$.
In this document, we adopt the ACB models, in particular ACB5 as seen in Table II of Ref.~\cite{Paschalidis2018}.
We begin in Sec.~\ref{sec:nuclear} by fitting new binary Love relations for nucleonic matter as was done in~\cite{Yagi:binLove}.
This is followed in Sec.~\ref{sec:hybrid} by an analysis and discussion into how well hybrid stars agree with the improved binary Love relations

@@

We have now shown that binary NS merger observations can help improve universal relations - the question is: is it worth it?
Current interferometer sensitivities are not yet small enough to accurately constrain $\tilde{\Lambda}$.
For example, GW170817 was detected by LIGO observing run 2 (``O2")\zc{cite} and was able to constrain $\tilde{\Lambda}$ to a $90\%$ confidence interval of 325 (this relates to a standard deviation of $\sigma_{\tilde{\Lambda}}$=198) centered at $\tilde{\Lambda}=395$.
This corresponds to statistical uncertainties on the order of $\sim 82\%$, which vastly dominates the error budget compared to the small systematic errors picked up by EoS variation in the universal relations, which can be seen in Fig.~\ref{fig:variances}.
This implies that currently, improved universal relations will not make any difference in tidal parameter extraction.
However, future detectors (for example aLIGO, a\texttt{++}, Voyager, CE, and ET) are planned with large reductions in sensitivity - both decreasing the statistical errors and allowing for a larger binary NS merger detection rate; further reducing uncertainties.
In Sec.~\ref{sec:observations}, we analyze this further and discuss when improved universal relations will make a difference.
\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{variances.png}
\end{center}
\caption{
\zc{Ask Katerina for this data. Convert $\Lambda_s$ to $\tilde{\Lambda}$, then redo this image in xmgrace. Remove posterior eoss, Add a gridline at the maximum errors for both constrained and unconstrained, and maybe even show the relative errors as well? Probably remove the means as well.}
Absolute difference from the fit on $\tilde{\Lambda}$ from EoS variation in the binary Love universal relations plotted with varying mass ratio $q$.
Observe that the maximum systematic error attained for the constrained EoSs is $\sigma_{\tilde{\Lambda}}=10$ (this is found to be $20$ for the unconstrained EoSs).
This is negligible compared to the statistical errors accrued on paramter extraction from GW170817.
}
\label{fig:variances}
\end{figure}

>>>>

\definecolor{red(ncs)}{rgb}{0.77, 0.01, 0.2}
\newcommand{\ny}[1]{\textcolor{blue}{NY: #1} }
\newcommand{\kc}[1]{\textcolor{red}{KC: #1} }

@@

In the current analysis, we consider only EoSs which have been constrained by the recent binary neutron star merger event, GW170817.
Important recent work~\cite{LIGO:posterior} uniformly sampled EoS parameters in order to derive a marginalized posterior on the pressure as a function of mass density, as seen in Fig. 2 of~\cite{LIGO:posterior}.
In our work, we similarly sample spectral coefficients within the ranges: $\gamma_0 \in \lbrack 0.2,2 \rbrack$, $\gamma_1 \in \lbrack -1.6,1.7 \rbrack$, $\gamma_2 \in \lbrack -0.6,0.6 \rbrack$, $\gamma_0 \in \lbrack -0.02,0.02 \rbrack$, and further restricting the adiabatic index to be $\Gamma \in \lbrack 0.6,4.5 \rbrack$, ensuring our parameterization exposes a wide range of viable EoSs~\cite{Lindblom:parameters}.
In addition, we impose similar constraints as was done in~\cite{LIGO:posterior}: (i) causality within 10\%, and (ii) EoS priors must support NS masses up to $1.97 \text{ M}_{\odot}$, consistent with astrophysical observations.
We then utilize the posterior defined by GW170817 by rejecting any EoS sample which does not lie within the 90\% credible posterior region, resulting in a sample of 35 EoSs shown in Fig.~\ref{fig:eos}.
\kc{Why aren't we directly using the gamma's that Carl provided?}
Following~\cite{Read2009}, the parameterized high-density core EoSs generated above are matched to the low-density crust EoS Sly~\cite{Douchin:2001sv} at about half of the nuclear saturation density, $\rho_{\text{stitch}}=1.3 \times 10^{14} \text{ g/cm}^3$.
For comparison, we also include a second sample of randomly generated EoSs, without restricting to GW170817s posterior. {\ny{It would be good to make this figure a double pane one, so that you can show the constrained set and the unconstrained set side-by-side.}}
\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{EoSs.eps}
\end{center}
\caption{
Representative set of EoSs used in our analysis. 
These were populated by uniformly sampling the spectral coefficient parameter space, then restricting to the 90\% credible level shown in cyan~\cite{LIGO:posterior}.
In addition, the following constraints were also applied: (i) the EoS must have a causal structure within 10\%, and (ii) the EoS must support maximum NS mass of at least $1.97 \text{ M}_{\odot}$, consistent with astrophysical observations.
Finally, the high density core EoSs generated above were stitched to the low-density crust EoS of SLy~\cite{Douchin:2001sv} at $\rho_{\text{stitch}}=1.3 \times 10^{14} \text{ g/cm}^3$.
}
\label{fig:eos}
\end{figure}

@@

\kc{why not study the I-Q relation too?}
\subsection{Binary love relations}\label{sec:binary}
Next we consider improvements to the binary Love universal relations.
In particular, we consider two distinct classes of NSs: nuclear matter EoSs and hybrid quark-hadron star EoSs~\cite{Paschalidis2018,Alford:2017qgh,1971SvA....15..347S,Zdunik:2012dj,Alford:2013aca}.
In the latter, the low-density nucleonic crust region of the EoS experiences a first-order phase transition into a high density quark matter core at a given transition pressure $P_{\text{tr}}$.
In this document, we adopt the ACB models, in particular ACB5 as seen in Table II of Ref.~\cite{Paschalidis2018}.
We begin in Sec.~\ref{sec:nuclear} by fitting new binary Love relations for nucleonic matter as was done in~\cite{Yagi:binLove}.
This is followed in Sec.~\ref{sec:hybrid} by an analysis and discussion into how well hybrid stars agree with the improved binary Love relations

@@

We have now shown that binary NS merger observations can help improve universal relations - the question is: is it worth it?
Current interferometer sensitivities are not yet small enough to accurately constrain $\tilde{\Lambda}$.
For example, GW170817 was detected by LIGO observing run 2 (``O2")\zc{cite} and was able to constrain $\tilde{\Lambda}$ to a $90\%$ confidence interval of 325 (this relates to a standard deviation of $\sigma_{\tilde{\Lambda}}$=198) centered at $\tilde{\Lambda}=395$.
This corresponds to statistical uncertainties on the order of $\sim 82\%$, which vastly dominates the error budget compared to the small systematic errors picked up by EoS variation in the universal relations, which can be seen in Fig.~\ref{fig:variances}.
This implies that currently, improved universal relations will not make any difference in tidal parameter extraction.
However, future detectors (for example aLIGO, a\texttt{++}, Voyager, CE, and ET) are planned with large reductions in sensitivity - both decreasing the statistical errors and allowing for a larger binary NS merger detection rate; further reducing uncertainties.
In Sec.~\ref{sec:observations}, we analyze this further and discuss when improved universal relations will make a difference.
\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{variances.png}
\end{center}
\caption{
\zc{Ask Katerina for this data. Convert $\Lambda_s$ to $\tilde{\Lambda}$, then redo this image in xmgrace. Remove posterior eoss, Add a gridline at the maximum errors for both constrained and unconstrained, and maybe even show the relative errors as well? Probably remove the means as well.}
Absolute difference from the fit on $\tilde{\Lambda}$ from EoS variation in the binary Love universal relations plotted with varying mass ratio $q$.
Observe that the maximum systematic error attained for the constrained EoSs is $\sigma_{\tilde{\Lambda}}=10$ (this is found to be $20$ for the unconstrained EoSs).
This is negligible compared to the statistical errors accrued on paramter extraction from GW170817.\kc{Similar to an earlier comment, I think we should use the "posterior" EoSs since they were directly computed from the data.}
}
\label{fig:variances}
\end{figure}
